Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-18 talk/pepm2011/paper.tex
add line numbers, try to gently rewrite things a bit

To optimize the trace, it is traversed from beginning to end.
Whenever a \lstinline{new} operation is seen, the operation is removed and a \emph{static
object}\footnote{Here ``static'' is meant in the sense of partial
evaluation, \ie known at partial evaluation time, not in the sense of static
allocation or static method.} is constructed and associated with the variable
that would have stored
the result of \lstinline{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \lstinline{set} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\lstinline{get} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\lstinline{guard_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type and thus the outcome of
the type check the guard does is statically known.

In the example from Section~\ref{sub:example}, the following operations in the upper half
of Figure~\ref{fig:unopt-trace} produce two
static objects, and can be completely removed from the optimized trace:

@@

The static object associated with $p_{5}$ would know that it is a
\lstinline{BoxedInteger} whose \lstinline{intval} field contains $i_{4}$; the
one associated with $p_{6}$ would know that it is a \lstinline{BoxedInteger}
whose \lstinline{intval} field contains the constant -100.

The subsequent operations in Figure~\ref{fig:unopt-trace},
 which use $p_{5}$ and $p_{6}$, could then be
optimized using that knowledge:

@@

So far we have only described what happens when static objects are used in guards and in
operations that read and write fields. When the static
object is used in any other operation, it cannot remain static. For example, when
a static object is stored in a globally accessible place, the object has to
be allocated, as it might live longer than one iteration of the loop
and because the partial evaluator looses track of it. This means that the static
object needs to be turned into a dynamic one, \ie lifted. This makes it
necessary to put operations into the residual code that allocate the
static object at runtime.

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
is hit. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well, recursively). This means that instead of a simple jump,
the following operations are emitted:

@@

\begin{figure}
\begin{lstlisting}[mathescape]
# arguments to the trace: $p_{0}$, $p_{1}$
guard_class($p_1$, BoxedInteger)
$i_2$ = get($p_1$, intval)
guard_class($p_0$, BoxedInteger)
$i_3$ = get($p_0$, intval)
$i_4$ = int_add($i_2$, $i_3$)
$i_9$ = int_add($i_4$, -100)

guard_class($p_0$, BoxedInteger)
$i_{12}$ = get($p_0$, intval)
$i_{14}$ = int_add($i_{12}$, -1)

$i_{17}$ = int_gt($i_{14}$, 0)
guard_true($i_{17}$)

$p_{15}$ = new(BoxedInteger)
set($p_{15}$, intval, $i_{14}$)
$p_{10}$ = new(BoxedInteger)
set($p_{10}$, intval, $i_9$)

@@

The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
The optimized trace contains only two allocations, instead of the original five,
and only three \lstinline{guard_class} operations, from the original seven.

>>>>

To optimize the trace, it is traversed from beginning to end. Every
optimization in the input trace is either removed, or new operations are
produced. Whenever a \lstinline{new} operation is seen, the operation it is
removed optimistically and a \emph{static object}\footnote{Here ``static'' is
meant in the sense of partial evaluation, \ie known at partial evaluation time,
not in the sense of static allocation or static method.} is constructed and
associated with the result variable. The static object describes the shape of
the original object, \ie where the values that would be stored in the fields of
the allocated object come from, as well as the type of the object.

When a \lstinline{set} that writes into a static object is optimized, the
corresponding shape description is updated and the operation is removed. This
means that the operation was done at partial evaluation time. When the
optimizer encounters a \lstinline{get} from a static object, the result is read
from the shape description, and the operation is also removed. Equivalently, a
\lstinline{guard_class} on a variable that has a shape description can be
removed as well, because the shape description stores the type and thus the
outcome of the type check the guard does is statically known. Operations that
have dynamic (\ie all other) objects as arguments are just left untouched by
the optimizer.

In the example from Section~\ref{sub:example}, the following operations
of Figure~\ref{fig:unopt-trace} (lines 10-17) produce two
static objects, and can be completely removed from the optimized trace:

@@

The static object associated with $p_{5}$ would store the knowledge that it is a
\lstinline{BoxedInteger} whose \lstinline{intval} field contains $i_{4}$; the
one associated with $p_{6}$ would store that it is a \lstinline{BoxedInteger}
whose \lstinline{intval} field contains the constant -100.

The subsequent operations (line 19-29) in Figure~\ref{fig:unopt-trace}, which
use $p_{5}$ and $p_{6}$, can then be optimized using that knowledge:

@@

So far we have only described what happens when static objects are used in guards and in
operations that read and write fields. When the static
object is used in any other operation, it cannot remain static. For example, when
a static object is stored in a globally accessible place, the object has to
be allocated, as it might live longer than one iteration of the loop and as
arbitrary \lstinline{set} operations could change it due to aliasing. This
means that the static
object needs to be turned into a dynamic one, \ie lifted. This makes it
necessary to put operations into the residual code that allocate the
static object at runtime.

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
is optimized. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well, recursively). This means that instead of a simple jump,
the following operations are emitted:

@@

\begin{figure}
\begin{lstlisting}[mathescape,numbers=right,escapechar=|,numberstyle = \tiny,numbersep=0pt, numberblanklines=false]
# arguments to the trace: $p_{0}$, $p_{1}$ |\setcounter{lstnumber}{2}|
guard_class($p_1$, BoxedInteger)           |\setcounter{lstnumber}{4}|
$i_2$ = get($p_1$, intval)
guard_class($p_0$, BoxedInteger)           |\setcounter{lstnumber}{7}|
$i_3$ = get($p_0$, intval)
$i_4$ = int_add($i_2$, $i_3$)              |\setcounter{lstnumber}{25}|
$i_9$ = int_add($i_4$, -100)               |\setcounter{lstnumber}{35}|

guard_class($p_0$, BoxedInteger)           |\setcounter{lstnumber}{38}|
$i_{12}$ = get($p_0$, intval)              |\setcounter{lstnumber}{42}|
$i_{14}$ = int_add($i_{12}$, -1)           |\setcounter{lstnumber}{50}|

$i_{17}$ = int_gt($i_{14}$, 0)             |\setcounter{lstnumber}{53}|
guard_true($i_{17}$)                       |\setcounter{lstnumber}{42}|

$p_{15}$ = new(BoxedInteger)               |\setcounter{lstnumber}{45}|
set($p_{15}$, intval, $i_{14}$)            |\setcounter{lstnumber}{26}|
$p_{10}$ = new(BoxedInteger)               |\setcounter{lstnumber}{28}|
set($p_{10}$, intval, $i_9$)               |\setcounter{lstnumber}{53}|

@@

The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
The optimized trace contains only two allocations, instead of the original five,
and only three \lstinline{guard_class} operations, from the original seven. The
line numbers are the lines where the operations occurred in the original trace
in Figure~\ref{fig:unopt-trace}.
